---
title: "Does Container Type Affect Perceived Soda Taste?"
subtitle: "A Randomized Complete Block Design Analysis"
author: "Bradley Petersen"
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: true
    code-summary: "Show code"
    code-tools: true
    theme: cosmo
    embed-resources: true
    output-file: index.html
execute:
  echo: true
  warning: false
  message: false
---

## Overview

This project examines whether the type of container used to serve soda is associated with differences in perceived **taste/crispness ratings**. The data contain three container treatments—**Glass Bottle**, **Can**, and **Plastic Cup**—observed across four blocks with three replicates per block-treatment combination.

The analysis uses a **randomized complete block design (RCBD)** framework to estimate the treatment effect while accounting for block-to-block variation. In addition to the primary ANOVA, the project includes exploratory visualization, descriptive statistics, model diagnostics, and Bonferroni-adjusted pairwise comparisons.

The data was collected by four experimenters, each of whom randomly selected three participants. Each participant received all three treatments in a randomly generated order.

### Research question

> After accounting for block effects, is there evidence that mean taste/crispness ratings differ among glass bottles, cans, and plastic cups?

### Statistical hypotheses

- **Null hypothesis:** Mean taste/crispness rating is the same for all three container treatments.
- **Alternative hypothesis:** At least one treatment mean differs.
- **Significance level:** $\alpha = 0.05$.

## Data

The dataset contains 36 observations: 4 blocks × 3 treatments × 3 replicates.

```{r}
#| label: create-data

soda_data <- read.csv("raw_soda_data.csv")

soda_data$block <- factor(soda_data$block)
soda_data$treatment <- factor(soda_data$treatment)
soda_data$replicate <- factor(soda_data$replicate)

```

```{r}
#| label: data-preview
#| echo: true

knitr::kable(
  head(soda_data, 10),
  caption = "First 10 observations in the dataset"
)
```

### Design balance

A balanced design is useful because each block-treatment combination contributes equally to the analysis.

```{r}
#| label: sample-sizes

sample_size_table <- table(soda_data$block, soda_data$treatment)
knitr::kable(sample_size_table, caption = "Observations per block-treatment combination")
```

## Exploratory analysis

### Descriptive statistics

```{r}
#| label: descriptive-statistics

mean_sd <- function(x) {
  c(n = length(x), mean = mean(x), sd = sd(x))
}

summary_by_treatment <- aggregate(
  taste_score ~ treatment,
  data = soda_data,
  FUN = mean_sd
)

summary_by_treatment <- do.call(data.frame, summary_by_treatment)
names(summary_by_treatment) <- c("Treatment", "n", "Mean", "SD")
summary_by_treatment$Mean <- round(summary_by_treatment$Mean, 2)
summary_by_treatment$SD <- round(summary_by_treatment$SD, 2)

knitr::kable(
  summary_by_treatment,
  caption = "Taste/crispness ratings by container treatment"
)
```

The **Can** treatment has the highest observed mean rating (`r round(mean(soda_data$taste_score[soda_data$treatment == "Can"]), 2)`), followed by **Glass Bottle** (`r round(mean(soda_data$taste_score[soda_data$treatment == "Glass Bottle"]), 2)`) and **Plastic Cup** (`r round(mean(soda_data$taste_score[soda_data$treatment == "Plastic Cup"]), 2)`). These descriptive differences do not by themselves establish a statistically significant treatment effect.

### Distribution by treatment

```{r}
#| label: fig-treatment-boxplot
#| fig-cap: "Distribution of taste/crispness ratings by container treatment. Black points indicate treatment means."
#| fig-width: 8
#| fig-height: 5

boxplot(
  taste_score ~ treatment,
  data = soda_data,
  xlab = "Container treatment",
  ylab = "Taste/crispness rating",
  ylim = c(0, 10)
)

points(
  seq_along(levels(soda_data$treatment)),
  tapply(soda_data$taste_score, soda_data$treatment, mean),
  pch = 19
)
```

### Treatment patterns across blocks

```{r}
#| label: fig-profile-plot
#| fig-cap: "Mean treatment ratings within each block."
#| fig-width: 8
#| fig-height: 5

interaction.plot(
  x.factor = soda_data$treatment,
  trace.factor = soda_data$block,
  response = soda_data$taste_score,
  fun = mean,
  type = "b",
  pch = 1:4,
  xlab = "Container treatment",
  ylab = "Mean taste/crispness rating",
  trace.label = "Block"
)
```

## Statistical model

The RCBD model is

$$Y_{ij} = \mu + \beta_i + \tau_j + \varepsilon_{ij},$$

where $\mu$ is the overall mean, $\beta_i$ is the effect of block $i$, $\tau_j$ is the effect of container treatment $j$, and $\varepsilon_{ij}$ represents random error.

```{r}
#| label: fit-model

rcbd_model <- aov(taste_score ~ block + treatment, data = soda_data)
anova_table <- anova(rcbd_model)

anova_display <- data.frame(
  Source = rownames(anova_table),
  Df = anova_table[, "Df"],
  `Sum Sq` = round(anova_table[, "Sum Sq"], 3),
  `Mean Sq` = round(anova_table[, "Mean Sq"], 3),
  `F value` = round(anova_table[, "F value"], 3),
  `p-value` = round(anova_table[, "Pr(>F)"], 4),
  check.names = FALSE
)

knitr::kable(anova_display, caption = "RCBD ANOVA results")
```

### Primary result

```{r}
#| label: extract-results
#| include: false

alpha <- 0.05
treatment_p <- anova_table["treatment", "Pr(>F)"]
treatment_f <- anova_table["treatment", "F value"]
treatment_df <- anova_table["treatment", "Df"]
error_df <- anova_table["Residuals", "Df"]
```

The treatment effect was **not statistically significant**, \$F(`r treatment_df`, `r error_df`) = `r round(treatment_f, 2)`, $p = `r round(treatment_p, 3)`$. At the 0.05 significance level, the analysis therefore fails to reject the null hypothesis.

In this dataset, there is **insufficient evidence that mean taste/crispness ratings differ by container type after accounting for block effects**. Although cans produced the highest observed mean rating, the difference was not large enough relative to within-treatment variability to support a statistically significant treatment effect.

## Model diagnostics

Model assumptions were evaluated using residual plots, a normal Q-Q plot, and a Bartlett test for equal variances across treatments.

```{r}
#| label: fig-model-diagnostics
#| fig-cap: "Standard diagnostic plots for the RCBD ANOVA model."
#| fig-width: 9
#| fig-height: 7

oldpar <- par(no.readonly = TRUE)
par(mfrow = c(2, 2))
plot(rcbd_model)
par(oldpar)
```

### Equal-variance check

```{r}
#| label: bartlett-test

bartlett_treatment <- bartlett.test(taste_score ~ treatment, data = soda_data)

bartlett_display <- data.frame(
  Statistic = round(unname(bartlett_treatment$statistic), 3),
  df = unname(bartlett_treatment$parameter),
  `p-value` = round(bartlett_treatment$p.value, 4),
  check.names = FALSE
)

knitr::kable(bartlett_display, caption = "Bartlett test for equal treatment variances")
```

With a Bartlett-test p-value of `r round(bartlett_treatment$p.value, 3)`, the data do not provide evidence of unequal treatment variances at the 0.05 level.

## Supplementary pairwise comparisons

Because the overall treatment ANOVA is not significant, pairwise comparisons are not the basis of the primary conclusion. They are included here to demonstrate the post-ANOVA workflow and the use of a Bonferroni correction for multiple comparisons.

```{r}
#| label: pairwise-comparisons

trt_levels <- levels(soda_data$treatment)
pairs <- t(combn(trt_levels, 2))

mse <- anova_table["Residuals", "Mean Sq"]
df_error <- anova_table["Residuals", "Df"]
trt_means <- tapply(soda_data$taste_score, soda_data$treatment, mean)
trt_n <- table(soda_data$treatment)

posthoc <- data.frame(
  Comparison = paste(pairs[, 1], "vs", pairs[, 2]),
  Mean_1 = as.numeric(trt_means[pairs[, 1]]),
  Mean_2 = as.numeric(trt_means[pairs[, 2]])
)

posthoc$Difference <- posthoc$Mean_1 - posthoc$Mean_2
posthoc$SE <- sqrt(
  mse * (
    1 / as.numeric(trt_n[pairs[, 1]]) +
      1 / as.numeric(trt_n[pairs[, 2]])
  )
)
posthoc$t_value <- posthoc$Difference / posthoc$SE
posthoc$p_raw <- 2 * pt(-abs(posthoc$t_value), df = df_error)
posthoc$p_bonferroni <- pmin(posthoc$p_raw * nrow(posthoc), 1)

tcrit_bonf <- qt(1 - alpha / (2 * nrow(posthoc)), df = df_error)
posthoc$CI_low <- posthoc$Difference - tcrit_bonf * posthoc$SE
posthoc$CI_high <- posthoc$Difference + tcrit_bonf * posthoc$SE

posthoc[, 2:ncol(posthoc)] <- lapply(posthoc[, 2:ncol(posthoc)], round, 3)
knitr::kable(posthoc, caption = "Bonferroni-adjusted pairwise treatment comparisons")
```

## Conclusion

This analysis found **no statistically significant evidence that container type affected perceived soda taste/crispness ratings** after accounting for block effects. The observed treatment means ranged from `r round(min(tapply(soda_data$taste_score, soda_data$treatment, mean)), 2)` to `r round(max(tapply(soda_data$taste_score, soda_data$treatment, mean)), 2)`, but the RCBD ANOVA indicated that these differences were consistent with random variability in the sample.

From a data-analysis perspective, this project demonstrates a complete experimental-analysis workflow: structuring a balanced dataset, performing exploratory analysis, fitting an RCBD ANOVA model, evaluating model assumptions, conducting multiple-comparison procedures, and communicating statistical results in a reproducible Quarto report.

## Reproducibility

This report is fully reproducible from the `.qmd` file. It uses only base R plus `knitr`, which is used by Quarto to render tables.

```{r}
#| include: false
#| label: session-info
#| echo: true

sessionInfo()
```
